Collision rates of planetesimals near mean-motion resonances

### 2020-11-30 main.tex
Finish calculation of collision + drift timescale for 1m bodies

\subsection{Collisions Between Planetesimals}

@@

\textbf{Although equation \ref{eq:saf} works well to describe the collision rate for a homogeneous collection of planetesimals, some problems arise when localized regions of commensurabilities are introduced. Namely, the resonances cause the orbits of planetesimals to precess, and an interface between secularly aligned and randomly oriented orbits arises on each side of the resonance. Additionally, an interface between dynamically cold and hot planetesimals forms at each side. For these reasons, an N-body treatment in which collisions are directly resolved is necessary to understand how the collision rate varies near the MMRs.}

\subsection{Gas Drag on Planetesimals}

\textbf{Over the course of many orbits, the residual gas from the primordial nebula can damp the eccentricities and inclinations of the planetesimals. For a planetesimal-sized body, the geometric cross section is much larger than the mean free path of the gas particles and the timescale to damp its eccentricity is given by \citep{1976PThPh..56.1756A}}

\textbf{\noindent where $m$, $e$ and $s$ are the mass, eccentricity and radius of the planetesimal. $C_{D}$ is a drag coefficient which is of order unity, $
\rho_{g}$ is the local density of the gas and $v_{th}$ is its thermal speed. At 3 au, the gaseous component of the solar nebula has a density of $4 
\times 10^{-10}$ g cm$^{-3}$ and a typical thermal speed of $10^{5}$ cm s$^{-1}$ \citep{1981PThPS..70...35H}. For a 100 km body with a 
density of 2 g cm$^{-3}$ and an eccentricity of 0.1, the damping timescale is on the order of $10^{5}$ years. This is much longer than the timescales related to secular forcing and libration near mean-motion resonances, as discussed above, but shorter than the typical protoplanetary disk lifetime and potentially the planetesimal formation timescale. For this reason, we do not model the effects of gas drag, although we consider its effects when constructing initial conditions for our simulations, which is discussed next.}

@@

\textbf{Although the features in figure \ref{fig:coll_hist_a} near the 3:1 and 2:1 MMRs appear qualitatively different, they can be explained by one single dynamical process. As discussed in section \ref{sec:saf}, the collision rate between planetesimals depends on both the encounter rate and the strength of gravitational focusing. The encounter rate grows as the relative velocity $v$ between the planetesimals increases, while gravitational focusing is most effective for bodies with small relative velocities. There is therefore an intermediate value of $v$ for which the average collision rate $\left< \sigma v \right>$ is at a minimum, which we will refer to as  $\left< \sigma v \right>_{0}$. The dynamical excitation induced by a mean motion resonance can either increase or decrease the local average collision rate, depending on the size and mass of the planetesimals, along with their unperturbed relative velocites.}

@@

\textbf{TODO: Introduce this a little mroe gently, start off by acknowledging the limitations of this model} To construct a radial dust profile from the collision statistics, we begin by making the assumption that any dust generated by collisions is strongly 
coupled to the gas. For a dust grain larger than the mean free path of the gas particles, the timescale to damp its eccentricity is given by \textbf{equation \ref{eq:t_edamp}}. At 3 au, the gaseous component of the solar nebula has a density of $4 
\times 10^{-10}$ g cm$^{-3}$ and a typical thermal speed of $10^{5}$ cm s$^{-1}$ \citep{1981PThPS..70...35H}. For a 1 mm dust grain with a 
density of 1 g cm$^{-3}$ and an eccentricity of 0.05, the damping timescale is $4 \times 10^{-3}$ years. This is orders of magnitude smaller than the 
orbital timescale, therefore we conclude that collisionally generated dust grains will immediately couple to the gas.

Operating under this assumption\footnote{As discussed in \cite{2017ApJ...850..103B}, the local generation of second-generation dust is only one 
consideration. The dust will also evolve spatially due to drag effects, but can also evolve in size or be re-accreted onto planetesimals. For simplicity, 
we focus on local dust sourcing only while acknowledging that subsequent evolution could affect the morphology.}, a map of the relative concentration 
of second-generation dust can be constructed from the cylindrical distance at which the collisions occur. \textbf{Here, cylindrical distance is measured as the distance between the central star and the collision in the ($r, \theta$) plane.} This is shown in figure \ref{fig:coll_hist_r}. 
Similar to figure \ref{fig:coll_hist_a}, we use a KDE with a width of 0.02 au to assemble the collision locations into a radial distribution. Most 
strikingly, the bump that was present near the 2:1 resonance is no longer visible in cylindrical distance space, although it suddenly appears again in 
the e3m2 simulation. The dip that is visible near the 3:1 MMR also becomes a bump in the highest eccentricity case. To determine how the resonant 
bodies actually contribute to the radial collision profile, we excluded collisions that fall between $2.495 < a < 2.505$ au (near the 3:1 resonance) and 
between $3.2 < a < 3.35$ au (near the 2:1 resonance), which is shown by the dashed curve. \textbf{As in figure \ref{fig:coll_hist_a}, the solid curves are normalized such that the area underneath is equal to 1. For the dashed curves, the normalization factor is scaled down according to the total number of collisions in the entire sample and the number of collisions in the subsample which excludes the collisions in resonance.} Qualitatively, none of the bump or dip features present 
are removed by making this exclusion. This suggests that the features near resonance seen in semimajor axis space become smeared out in 
cylindrical distance space due to the large spread in eccentricity and orbital orientation.

>>>>

\subsection{Collisions Between Planetesimals}\label{sec:colleq}

@@

\textbf{Although equation \ref{eq:saf} works well to describe the collision rate for a homogeneous collection of planetesimals, some problems arise when localized regions of commensurabilities are introduced. Namely, the resonances cause the orbits of planetesimals to precess, and an interface between secularly aligned and randomly oriented orbits arises on each side of the resonance. Additionally, an interface between dynamically cold and hot planetesimals forms at each side. All of these effects cause the number density, collision cross section and encounter velocity to spatially vary in a rather abrupt manner. For these reasons, an N-body treatment in which collisions are directly resolved is necessary to understand how the collision rate varies near the MMRs.}

\subsection{Gas Drag on Planetesimals}\label{sec:pl_drag}

\textbf{Over the course of many orbits, the residual gas from the primordial nebula can damp the eccentricities and inclinations of the planetesimals. For a planetesimal-sized body, the geometric cross section is much larger than the mean free path of the gas particles and the timescale for gas drag to modify its orbit is given by \citep{1976PThPh..56.1756A}}

\begin{equation}\label{eq:t_edamp}
    t_{e, damp} = \frac{2 m_{pl}}{C_{D} \pi s_{pl}^{2} \rho_{g} v_{g}},
\end{equation}

\textbf{\noindent where $m_{pl}$ and $s_{pl}$ are the mass and radius of the planetesimal. $C_{D}$ is a drag coefficient which is of order unity, $
\rho_{g}$ is the local density of the gas and $v_{g}$ is the headwind velocity of the gas experienced by the planetesimal. At 3 au, the gaseous component of the solar nebula has a density of $4 
\times 10^{-10}$ g cm$^{-3}$ and the typical headwind experienced by a body on a Keplerian orbit is $\sim 5000$ cm s$^{-1}$ \citep{1981PThPS..70...35H}. For a 100 km body with a 
density of 2 g cm$^{-3}$ and an eccentricity of 0.1, the stopping timescale is on the order of $10^{6}$ years. This is much longer than the timescales related to secular forcing and libration near mean-motion resonances, as discussed above, but shorter than the typical protoplanetary disk lifetime and potentially the planetesimal formation timescale. For this reason, we do not model the effects of gas drag, although we consider its effects when constructing initial conditions for our simulations, which is discussed next.}

@@

\textbf{Although the features in figure \ref{fig:coll_hist_a} near the 3:1 and 2:1 MMRs appear qualitatively different, they can be explained by one single dynamical process. As discussed in section \ref{sec:colleq}, the collision rate between planetesimals depends on both the encounter rate and the strength of gravitational focusing. The encounter rate grows as the relative velocity $v$ between the planetesimals increases, while gravitational focusing is most effective for bodies with small relative velocities. There is therefore an intermediate value of $v$ for which the average collision rate $\left< \sigma v \right>$ is at a minimum, which we will refer to as  $\left< \sigma v \right>_{0}$. The dynamical excitation induced by a mean motion resonance can either increase or decrease the local average collision rate, depending on the size and mass of the planetesimals, along with their unperturbed relative velocites.}

@@

To construct a radial dust profile from the collision statistics, we begin by making \textbf{two assumptions: 1. any dust generated by collisions is strongly coupled to the gas and 2. any subsequent spatial evolution of the collisional debris is insignificant.}\textbf{So long as both of these assumptions are true, we can use the locations of planetesimal collisions in the plane of the disk to generate a dust profile for each simulation.}

\textbf{We first provide a justification for assumption (1), which is as follows:} For a dust grain larger than the mean free path of the gas particles, the timescale to damp its eccentricity is given by \citep{1976PThPh..56.1756A}

\noindent where $m$, $e$ and $s$ are the mass, eccentricity and radius of the dust grain. $C_{D}$ is a drag coefficient which is of order unity, $
\rho_{g}$ is the local density of the gas and $v_{th}$ is its thermal speed.\textbf{Using the same values for the properties of the solar nebula as we did in section \ref{sec:pl_drag} and assuming} a 1 mm dust grain with a 
density of 1 g cm$^{-3}$ and an eccentricity of 0.05, the damping timescale is $4 \times 10^{-3}$ years. This is orders of magnitude smaller than the 
orbital timescale, therefore we conclude that collisionally generated dust grains will immediately couple to the gas.

\textbf{Assumption (2) is less straightforward to justify and may not be reasonable in all cases. Although the spatial redistribution of the full-sized planetesimals and mm-sized dust grains is insignificant, the radial drift forces experienced by intermediate-sized bodies is much greater. Collisions between planetesimals do not immediately convert the mass to dust grains. Rather, the debris will undergo a collisional cascade process, which gradually grinds the resulting bodies down to smaller and smaller sizes. A proper treatment of the collisional cascade process, unfortunately, is quite complicated and is well beyond the scope of this work. However, one can safely ignore this process if the peak radial drift timescale is longer than the timescale for collisions between bodies that experience maximal radial drift.}

\textbf{In the inner part of the solar nebula, radial drift is maximized for meter-sized bodies. Following our same calculation in \ref{sec:pl_drag} with a 1 m body, we find that an object of this size will drift across the 2:1 MMR on a timescale of $\sim$ 300 years. Although we do not model collisions between bodies of this size, an upper limit on the collision timescale for 1 m bodies can be obtained by extrapolating the planetesimal collision rate measured in our simulations down to smaller-sized objects. For the e2m2 simulation, we find that 125 collisions occur withing the libration width of the 2:1 MMR during the 3,000 time window that collisions are recorded. There are 15,000 total planetesimals within this region, which means that a given planetesimal undergoes a collision every $\sim$ 370,000 years on average. If the entire planetesimal population in this region were converted to 1 meter bodies of the same bulk density, the collision timescale would decrease to $\sim$ 1 year. The number density of bodies increases by a factor of $10^{15}$, while the geometric collision cross section decreases by a factor of $10^{10}$. Additionally, the 1 meter bodies will not experience gravitational focusing like the planetesimals, but this only alters the collision timescale by a factor of order unity. From this crude calculation, we conclude that the collision timescale for bodies that experience maximal radial drift is smaller than their drift timescale. Therefore, our descision to ignore the spatial redistribution of the collisional debris is not entirely unfounded, although we would like to emphazise that the collision timescale estimated above is strictly a lower limit.}

Operating under \textbf{these assumptions} \footnote{As discussed in \cite{2017ApJ...850..103B}, the local generation of second-generation dust is only one 
consideration. The dust will also evolve spatially due to drag effects, but can also evolve in size or be re-accreted onto planetesimals. For simplicity, 
we focus on local dust sourcing only while acknowledging that subsequent evolution could affect the morphology.}, a map of the relative concentration 
of second-generation dust can be constructed from the cylindrical distance at which the collisions occur. \textbf{Here, cylindrical distance is measured as the distance between the central star and the collision in the ($r, \theta$) plane.} This is shown in figure \ref{fig:coll_hist_r}. 
Similar to figure \ref{fig:coll_hist_a}, we use a KDE with a width of 0.02 au to assemble the collision locations into a radial distribution. Most 
strikingly, the bump that was present near the 2:1 resonance is no longer visible in cylindrical distance space, although it suddenly appears again in 
the e3m2 simulation. The dip that is visible near the 3:1 MMR also becomes a bump in the highest eccentricity case. To determine how the resonant 
bodies actually contribute to the radial collision profile, we excluded collisions that fall between $2.495 < a < 2.505$ au (near the 3:1 resonance) and 
between $3.2 < a < 3.35$ au (near the 2:1 resonance), which is shown by the dashed curve. \textbf{As in figure \ref{fig:coll_hist_a}, the solid curves are normalized such that the area underneath is equal to 1. For the dashed curves, the normalization factor is scaled down according to the total number of collisions in the entire sample and the number of collisions in the subsample which excludes the collisions in resonance.} Qualitatively, none of the bump or dip features present 
are removed by making this exclusion. This suggests that the features near resonance seen in semimajor axis space become smeared out in 
cylindrical distance space due to the large spread in eccentricity and orbital orientation.
